Batch Effects in the 1000 Genomes Project Data Confounds Modern Genomics Studies

### 2018-09-13 legacy.tex
modified model specs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Legacy Data Confounds Modern Genomics Studies}

@@

\subsection{Reference Cohorts}

The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
The International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses. 
Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.

@@

\subsection{Motivations}

@@

@@

The results section is organized as follows.
First, we describe the observed discrepancies within and between studies. We then identify the technical covariates that strongly correlate with the discordant values, and describe the distribution of such covariates across phases and populations of the 1kGP.    
Next, we identify lists of SNPs from the 1kGP whose validity is highly debatable, and show that major imputation servers impute the dubious SNPs.  
Even though these SNPs are likely technical artifacts, two recent studies report such SNPs as genome-wide significant or nearly significant hits.   
Fortunately, the main conclusions of the affected studies remain supported by other data. 
Yet this begs the question: When should we retire legacy reference data?

@@

			\section{Results}
	\subsection{A peculiar mutational signature in Japan}

@@

\subsection{Overlap of significant SNPs}
The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{combined analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.  
\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}

@@

@@

\section{Supplementary Figures}
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\title{Legacy Data Confounds Modern Genomics Studies}
\title{Batch Effects in the 1000 Genomes Project Data Confounds Modern Genomics Studies}

@@

\subsection{Batch Effects in Aging Reference Cohort Data}

The last 5 years have seen a drastic increase in the amount and quality of human genome sequence data. 
Reference cohorts such as the International HapMap Project \citep{HapMap2005}, the 1000 Genomes Project \citep{1000GenomesProjectConsortium2010,The1000GenomesProjectConsortium2012}, and the Simons Diversity project \citep{Mallick2016}, for example, have made thousands of genome sequences publicly available for population and medical genetic analyses. 
Many more genomes are available indirectly through servers providing imputation services \citep{HapMap2005} or summary statistics for variant frequency estimation \citep{Lek2016}.

@@

\subsection{Mutational Signatures}

@@

 \subsection{Irreproducibility of a Japanese Mutational Signature}

@@

@@

			\section{Results}
The results section is organized as follows.
First, we describe the observed discrepancies within and between studies. We then identify the technical covariates that strongly correlate with the discordant values, and describe the distribution of such covariates across phases and populations of the 1kGP.    
Next, we identify lists of SNPs from the 1kGP whose validity is highly debatable, and show that major imputation servers impute the dubious SNPs.  
Even though these SNPs are likely technical artifacts, two recent studies report such SNPs as genome-wide significant or nearly significant hits.   
Fortunately, the main conclusions of the affected studies remain supported by other data. 
Yet this begs the question: When should we retire legacy reference data?

	\subsection{A peculiar mutational signature in Japan}

@@

\subsection{Overlap of significant SNPs}
%Define the approach before calling a genome wide simple logistic regression analysis
The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}), and 3990 SNPs were statistically significant after a \luke{combined analysis} (taking the sum of the deviations from each population and adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant), as shown on Figure \ref{Manhattan}.
%Wanted a test statistic to represent the association across all populations, to do that we summed across the deviations for each population. 
%
%\sgcomment{Look at the frequency distributions of these SNPs in GnomAD}

@@

\subsection{Model Specification}

${logit}E(Y_{si}) = \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{i} $

for $i = 1,\hdots, n_p$

where ${Q}$ is the average quality of mapped bases, ${i}$ is the individual, ${s}$ is SNP, ${p}$ is the population.

@@

\clearpage
\section{Supplementary Figures}

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
